Synthesis report

### 2020-02-20 analysis/reports/synthesis.Rmd
render on hpg

all_di <- all_di %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)),
         log_nsamples = log(nsamples)) %>%
  mutate(prop_found = exp(log_nsamples - log_nparts))
```

@@

```{r datasets in s and n space}
ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = dat)) +
  geom_point(alpha = .8) +
  theme_bw() +
  scale_color_viridis_d()

@@

```{r size of fs}
ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = log_nparts)) +
  geom_point(alpha = .5) +
  theme_bw() +
  scale_color_viridis_c(option = "magma", end = .9, begin = .1)

@@

Here is how the size of the feasible set maps on to N/S:

```{r nparts vs avgn}
ggplot(filter(all_di, singletons == F), aes(x = log(n0 / s0), y = log_nparts, color = log(s0))) +
  theme_bw() +
  geom_point(alpha = .2) +
  scale_color_viridis_c()

@@

```{r nsamples}
ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = log_nsamples)) +
  geom_point(alpha = .3) +
  theme_bw() +
  scale_color_viridis_c(option = "magma", end = .9)

ggplot(filter(all_di, singletons == F, nsamples <  max(nsamples)), aes(x = log(s0), y = log(n0), color = log_nsamples)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(option = "magma", end = .9)
```

Only in the very smallest communities do we get appreciably fewer than `r max(all_di$nsamples)` samples.

@@

```{r nsamples vs nparts}
ggplot(filter(all_di, singletons == F), aes(x = prop_found)) +
  theme_bw() +
  geom_histogram(bins = 100)

ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = prop_found)) +
  geom_point(alpha = .7) +
  geom_point(data = filter(all_di, singletons == F, prop_found == 1), color = "grey") +
  theme_bw() +
  scale_color_viridis_c(option = "magma", end = .9, direction = -1) +
  theme(legend.position = "top")

@@

For about `r 100 * mean(filter(all_di, singletons == F)$prop_found == 1)`% of sites, we found all the elements of the FS. The *vast majority* of this is FIA.

#

>>>>

all_di <- all_di %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)),
         log_nsamples = log(nsamples)) %>%
  mutate(prop_found = exp(log_nsamples - log_nparts)) %>%
  mutate(found_all = prop_found==1)
```

@@

```{r datasets in s and n space}
ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = dat)) +
  geom_point(alpha = .8) +
  theme_bw() +
  theme(legend.position = "top") + scale_color_viridis_d()

@@

```{r size of fs}
ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = log_nparts)) +
  geom_point(alpha = .5) +
  theme_bw() +
  theme(legend.position = "top") + scale_color_viridis_c(option = "magma", end = .9, begin = .1)

@@

Here is how the size of the feasible set maps on to N/S. It increases with n0/s0 and s0.

```{r nparts vs avgn}
ggplot(filter(all_di, singletons == F), aes(x = log(n0 / s0), y = log_nparts, color = log(s0))) +
  theme_bw() +
  geom_point(alpha = .2) +
  theme(legend.position = "top") + scale_color_viridis_c()

@@

```{r nsamples}
ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = log_nsamples)) +
  geom_point(alpha = .3) +
  theme_bw() +
  theme(legend.position = "top") + scale_color_viridis_c(option = "magma", end = .9)

Only in small communities do we get appreciably fewer than `r max(all_di$nsamples)` samples.

@@

```{r nsamples vs nparts, fig.dim = c(4,3)}
ggplot(filter(all_di, singletons == F), aes(x = prop_found)) +
  theme_bw() +
  geom_histogram(bins = 100)

ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = prop_found)) +
  geom_point(alpha = .7) +
  geom_point(data = filter(all_di, singletons == F, prop_found == 1), color = "grey") +
  theme_bw() +
  scale_color_viridis_c(option = "magma", end = .9, direction = -1) #+
#theme(legend.position = "top")

@@

For about `r 100 * mean(filter(all_di, singletons == F)$prop_found == 1)`% of sites, we found all the elements of the FS. The *vast majority* of this is FIA - here is what happens if we take out FIA:

```{r prop found no fia, fig.dim = c(4,3)}
ggplot(filter(all_di, singletons == F, dat != "fia_short"), aes(x = prop_found)) +
  theme_bw() +
  geom_histogram(bins = 100)

ggplot(filter(all_di, singletons == F, dat != "fia_short"), aes(x = log(s0), y = log(n0), color = prop_found)) +
  geom_point(alpha = .7) +
  geom_point(data = filter(all_di, singletons == F, dat != "fia_short", prop_found == 1), color = "grey") +
  theme_bw() +
  scale_color_viridis_c(option = "magma", end = .9, direction = -1)# +
#theme(legend.position = "top")

ggplot(filter(all_di, singletons == F, dat != "fia_short"), aes(x = prop_found, y =nsamples)) +
  theme_bw() +
  geom_vline(xintercept = c((.95), (.1))) +
  geom_point(alpha = .3) +
  geom_point(data = filter(all_di, singletons == F, prop_found == 1, dat != "fia_short"), color = "grey")

prop_found_ecdf <- ecdf(filter(all_di,singletons == F, dat != "fia_short")$prop_found)

Without FIA, we find all the samples about `r 100 * mean(filter(all_di, singletons == F, dat != "fia_short")$prop_found == 1)`% of the time.

# Distribution of percentile values

## Skewness

Here is the overall distribution of skewness, and if we split based on whether we found all the samples:
```{r skew percentiles}
ggplot(filter(all_di, singletons == F), aes(x = skew_percentile, y =..count.. / sum(..count..))) +
  geom_histogram(bins = 100) +
  theme_bw() +
  geom_hline(yintercept = .01)
ggplot(filter(all_di, singletons == F), aes(x = skew_percentile)) +
  theme_bw() +
  geom_histogram(bins = 100) +
  facet_wrap(vars(found_all), scales = "free_y")
```

When we found all the samples, the percentiles are more evenly distributed. I do not read much into the spike at 0 for those communities, because skewness is bizarre for very small communities.

Here is how skewness maps with S and N:
```{r skew v s and n, fig.dim = c(10,10)}

ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = skew_percentile)) +
  geom_point(alpha = .5) +
  theme_bw() +
  theme(legend.position = "top") + scale_color_viridis_c(option = "plasma", end = .9)

# ggplot(filter(all_di, singletons == F), aes(x = log(n0/s0), y = skew_percentile, color = prop_found)) +
#   geom_point() +
#   theme_bw() +
#   scale_color_viridis_c(option = "plasma", end = .9, direction = -1) +
#   theme(legend.position = "top")
# 
# ggplot(filter(all_di, singletons == F), aes(x = log(n0 / s0), y = skew_percentile, color = log(s0))) +
#   geom_point(alpha = .2) +
#   theme_bw() +
#   theme(legend.position = "top") + scale_color_viridis_c()
```

The very low skewness values are down in the very small and very weird communities. There may be variation along S and N elsewhere, but it is hard to parse.

## Simpson

Here is the overall evenness distribution, and split by whether we found 'em all:

```{r simpson percentiles}
ggplot(filter(all_di, singletons == F), aes(x = simpson_percentile, y =..count.. / sum(..count..))) +
  geom_histogram(bins = 100) +
  theme_bw() +
  geom_hline(yintercept = .01)

ggplot(filter(all_di, singletons == F), aes(x = simpson_percentile)) +
  theme_bw() +
  geom_histogram(bins = 100) +
  facet_wrap(vars(found_all))

Simpson is less evenly distributed than skewness. Again, where we found them all, we don't see the disproportionately common low percentile values.

Here is how Simpson behaves in S and N space:

```{r even v s and n, fig.dim = c(10,10)}
# ggplot(filter(all_di, singletons == F), aes(x = log(n0 / s0), y = simpson_percentile, color = log(s0))) +
#   geom_point(alpha = .2) +
#   theme_bw() +
#   theme(legend.position = "top") + scale_color_viridis_c()

ggplot(filter(all_di, singletons == F), aes(x = log(s0), y = log(n0), color = simpson_percentile)) +
  geom_point(alpha = .5) +
  theme_bw() +
  theme(legend.position = "top") + scale_color_viridis_c(option = "plasma", end = .9)
# ggplot(filter(all_di, singletons == F), aes(x = log(s0/n0), y = simpson_percentile, color = prop_found)) +
#   geom_point() +
#   theme_bw() +
#   scale_color_viridis_c(option = "plasma", end = .9, direction = -1) +
#   theme(legend.position = "top")
```

There is unusual behavior where S is large and N/S is relatively small (log N/S <= 1.5), where evenness is unusually *high*.

For both skew and evenness, we do not see non-extreme percentile values in large communities:

```{r non extremes}

ggplot(filter(all_di, singletons == F), aes(x = log(n0 / s0), y = skew_percentile, color = log(n0))) +
  geom_point(alpha = .1) +
  theme_bw() +
  theme(legend.position = "top") + 
  scale_color_viridis_c(end = .9)

ggplot(filter(all_di, singletons == F), aes(x = log(n0 / s0), y = simpson_percentile, color = log(n0))) +
  geom_point(alpha = .1) +
  theme_bw() +
  theme(legend.position = "top") + 
  scale_color_viridis_c(end = .9)

## Singletons

Here is how singletons change percentiles, broken out by whether or not we found all the samples:

```{r singletons overall}

singletons_effect <- all_di %>%
  select(dat, nsamples, site, singletons, skew_percentile, simpson_percentile, found_all) %>%
  tidyr::pivot_wider(names_from = singletons, values_from = c(skew_percentile, simpson_percentile, nsamples, found_all)) %>%
  mutate(skewdiff = skew_percentile_TRUE - skew_percentile_FALSE,
         simpdiff= simpson_percentile_TRUE - simpson_percentile_FALSE,
         found_all = (found_all_TRUE + found_all_FALSE) == 2)

ggplot(data = singletons_effect, aes(x = skew_percentile_FALSE, y = skew_percentile_TRUE)) +
  geom_point(alpha = .1) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, color = "green") +
  facet_wrap(vars(found_all))

ggplot(data = singletons_effect, aes(x = simpson_percentile_FALSE, y = simpson_percentile_TRUE)) +
  geom_point(alpha = .1) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, color = "green") +
  facet_wrap(vars(found_all))

The rarefaction-inflated datasets are strongly // the raw vectors. They have more extreme skewness and evenness values, relative to their feasible sets, than the raw vectors. This is almost always true for evenness, with a little more noise in the skewness signal. But either way, very strong.

# Manipulations
# MACD

```{r load macd}
cache_loc = "macdb"

## Set up the cache and config
db <- DBI::dbConnect(RSQLite::SQLite(), here::here("analysis", "drake", paste0("drake-cache-", cache_loc, ".sqlite")))
cache <- storr::storr_dbi("datatable", "keystable", db)

all_di_macd <- readd(all_di_manip, cache = cache) 
macd_dat <- readd(dat_s_dat_macdb, cache = cache)

DBI::dbDisconnect(db)
rm(cache)

macd_dat <- macd_dat %>%
  group_by(site, dat, singletons, sim, source) %>%
  summarize(s0 = max(rank),
            n0 = sum(abund)) %>%
  ungroup()

all_di_macd <- left_join(all_di_macd, macd_dat, by = c("site", "dat", "singletons", "sim", "source")) %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)),
         log_nsamples = log(nsamples)) %>%
  mutate(prop_found = exp(log_nsamples - log_nparts)) %>%
  mutate(found_all = prop_found==1)

ggplot(data = all_di, aes(x = log(s0), y = log(n0))) + geom_point(alpha = .1) + geom_point(data = all_di_macd, aes(x = log(s0), y = log(n0)), color = "blue", alpha = .9) + theme_bw()

ggplot(data = filter(all_di, !singletons), aes(x = log(s0), y = log(n0))) + geom_point(alpha = .1) + geom_point(data = filter(all_di_macd, !singletons), aes(x = log(s0), y = log(n0)), color = "blue", alpha = .9) + theme_bw()

Here are the distributions of skew and evenness, overall.

```{r macd overall}

ggplot(data = all_di_macd, aes(x = skew_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(found_all), scales = "free_y")

ggplot(data = all_di_macd, aes(x = simpson_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(found_all), scales = "free_y")
```

Here is how manipulation affects things:

```{r macd ctrlcomp, fig.dim = c(4,4)}

macd_comparisons <- read.csv(here::here("working-data", "macdb_data", "orderedcomparisons.csv"), header = F, stringsAsFactors = F)

colnames(macd_comparisons) <- c("studyID", "control", "site")
macd_comparisons <- macd_comparisons %>%
  mutate(site = as.character(site), control = as.character(control))

cc_di <- all_di_macd %>%
  filter(singletons == FALSE, treatment == "comparison") %>%
  left_join(macd_comparisons, by = c("studyID", "site")) %>%
  select(dat, site, skew_percentile, simpson_percentile, studyID, control) %>%
  rename(comparison = site) %>%
  mutate(ctrl_skew_percentile = NA,
         ctrl_simpson_percentile = NA)

for(i in 1:nrow(cc_di)) {

  control_row <- filter(all_di_macd, site == cc_di$control[i], singletons == F) %>%
    unique()
  if(nrow(control_row) >0 ) {
    cc_di$ctrl_simpson_percentile[i] <- control_row$simpson_percentile
    cc_di$ctrl_skew_percentile[i] <- control_row$skew_percentile
  }
}

cc_di <- cc_di %>%
  mutate(simpson_change = simpson_percentile - ctrl_simpson_percentile,
         skew_change = skew_percentile - ctrl_skew_percentile)

ggplot(data = cc_di, aes(x = skew_percentile, y = ctrl_skew_percentile)) +
  geom_point(alpha = .5) +
  #  xlim(0, 100) +
  # ylim(0, 100) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, color = "green")

ggplot(data = cc_di, aes(x = simpson_percentile, y = ctrl_skew_percentile)) +
  geom_point(alpha = .5) +
  xlim(0, 100) +
  ylim(0, 100) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, color = "green")

ggplot(data = cc_di, aes(x = simpson_change)) +
  geom_histogram() +
  theme_bw()

ggplot(data = cc_di, aes(x = skew_change)) +
  geom_histogram() +
  theme_bw()

ggplot(data = cc_di, aes(x = ctrl_simpson_percentile, y = simpson_change)) +
  geom_point(alpha = .3) +
  geom_hline(yintercept = 0) +
  theme_bw()

ggplot(data = cc_di, aes(x = ctrl_skew_percentile, y = skew_change)) +
  geom_point(alpha = .3) +
  geom_hline(yintercept = 0) +
  theme_bw()

print(t.test(cc_di$skew_percentile, cc_di$ctrl_skew_percentile, paired = T)
)
print(t.test(cc_di$simpson_percentile, cc_di$ctrl_simpson_percentile, paired = T)
)

print(wilcox.test(cc_di$skew_percentile, cc_di$ctrl_skew_percentile, paired = T))
print(wilcox.test(cc_di$simpson_percentile, cc_di$ctrl_simpson_percentile, paired = T))

Change is going to be bounded at 100 and 0: you can't go up or down from there. (Another argument for increasing the number of samples?)

# Portal plant manips

```{r load pp}
cache_loc = "portal_plants_manip"

## Set up the cache and config
db <- DBI::dbConnect(RSQLite::SQLite(), here::here("analysis", "drake", paste0("drake-cache-", cache_loc, ".sqlite")))
cache <- storr::storr_dbi("datatable", "keystable", db)

all_di_p<- readd(all_di_manip, cache = cache)

all_di_p <- all_di_p %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)),
         log_nsamples = log(nsamples)) %>%
  mutate(prop_found = exp(log_nsamples - log_nparts)) %>%
  mutate(found_all = prop_found==1)

DBI::dbDisconnect(db)
rm(cache)
```

Nsamples, singletons

```{r pp overall}
# 
# ggplot(data = all_di_p, aes(x = nsamples)) +
#   geom_histogram(bins = 100) +
#   theme_bw() +
#   geom_vline(xintercept = 2000)
# 
# all_di_p <- filter(all_di_p, nsamples >= 2000)

ggplot(data = all_di_p, aes(x = skew_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(found_all), scales = "free_y")

ggplot(data = all_di_p, aes(x = simpson_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(found_all), scales = "free_y")
# 
# pp_singletons_effect <- all_di_p %>%
#   select(singletons, skew_percentile, simpson_percentile, year, plot, treatment, season) %>%
#   tidyr::pivot_wider(names_from = singletons, values_from = c("skew_percentile", "simpson_percentile"))
# 
# ggplot(data = pp_singletons_effect, aes(x = skew_percentile_FALSE, y  = skew_percentile_TRUE)) +
#   geom_point(alpha = .1) +
#   geom_abline(intercept = 0, slope = 1, color = "green") +
#   theme_bw()
# 
# 
# ggplot(data = pp_singletons_effect, aes(x = simpson_percentile_FALSE, y  = simpson_percentile_TRUE)) +
#   geom_point(alpha = .1) +
#   geom_abline(intercept = 0, slope = 1, color = "green") +
#   theme_bw()
```

By treatment, season

```{r portal trtmt}
ggplot(data = filter(all_di_p, singletons == F), aes(x = skew_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(season, treatment), scales = "free_y")

ggplot(data = filter(all_di_p, singletons == F), aes(x = treatment, y = skew_percentile)) +
  geom_boxplot() +
  theme_bw() +
  facet_wrap(vars(season))

ggplot(data = filter(all_di_p, singletons == F), aes(x = simpson_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(season, treatment), scales = "free_y")

ggplot(data = filter(all_di_p, singletons == F), aes(x = treatment, y = simpson_percentile)) +
  geom_boxplot() +
  theme_bw() +
  facet_wrap(vars(season))
```

<!-- By year -->

<!-- ```{r plants year, fig.height = 25, fig.width = 5} -->

<!-- ggplot(data = filter(all_di_p, singletons == F), aes(x = treatment, y = skew_percentile)) + -->
<!--   geom_boxplot() + -->
<!--   theme_bw() + -->
<!--   facet_grid(rows = vars(year), cols = vars(season), scales = "free_y") -->
<!-- ggplot(data = filter(all_di_p, singletons == F), aes(x = treatment, y = simpson_percentile)) + -->
<!--   geom_boxplot() + -->
<!--   theme_bw() + -->
<!--   facet_grid(rows = vars(year), cols = vars(season), scales = "free_y") -->
<!-- ``` -->

Median
```{r pp median, fig.width = 12, fig.height = 12}

pp_median <- all_di_p %>%
  filter(singletons == F) %>%
  group_by(season, year, treatment) %>%
  summarize(skew_percentile = median(skew_percentile),
            simpson_percentile = median(simpson_percentile)) %>%
  ungroup() %>%
  tidyr::pivot_wider(names_from = treatment, values_from = c(skew_percentile, simpson_percentile))

sk_x <- ggplot(data = pp_median, aes(x = skew_percentile_control, y = skew_percentile_exclosure)) +
  geom_point(alpha = .5) +
  facet_wrap(vars(season)) +
  theme_bw() +
  ggtitle("Skew Ctrl v Exclosure") +
  geom_abline(intercept = 0, slope = 1, color = "green") +
  xlim(40, 100) +
  ylim(40, 100)

sk_r <- ggplot(data = pp_median, aes(x = skew_percentile_control, y = skew_percentile_removal)) +
  geom_point(alpha = .5) +
  facet_wrap(vars(season)) +
  theme_bw() +
  ggtitle("Skew Ctrl v Removal") +
  geom_abline(intercept = 0, slope = 1, color = "green") +
  xlim(40, 100) +
  ylim(40, 100)

gridExtra::grid.arrange(grobs = list(sk_x, sk_r), nrow = 2)

sp_x <- ggplot(data = pp_median, aes(x = simpson_percentile_control, y = simpson_percentile_exclosure)) +
  geom_point(alpha = .5) +
  facet_wrap(vars(season)) +
  theme_bw() +
  ggtitle("Simpson Ctrl v Exclosure") +
  geom_abline(intercept = 0, slope = 1, color = "green") +
  xlim(0, 40) +
  ylim(0, 40)

sp_r <- ggplot(data = pp_median, aes(x = simpson_percentile_control, y = simpson_percentile_removal)) +
  geom_point(alpha = .5) +
  facet_wrap(vars(season)) +
  theme_bw() +
  ggtitle("Simpson Ctrl v Removal") +
  geom_abline(intercept = 0, slope = 1, color = "green") +
  xlim(0, 40) +
  ylim(0, 40)

gridExtra::grid.arrange(grobs = list(sp_x, sp_r), nrow = 2)
```
